---
title: "Obligatory Exercise 2"
subtitle: "TMA4275 Lifetime analysis Spring 2019 "
author: "Candidate number: 10006"
date: "23 mars 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(survival)
library(survminer)
library(kableExtra)
library(SurvRegCensCov)
library(rms)
```
In this exercise we used the \textbf{R}-libraries:
```{r, eval=FALSE}
# for ggplot and dataframe
library(tidyverse)
# for survival analysis functions
library(survival)
# to plot survival curves in ggplot
library(survminer)
# to create tables in rmd
library(kableExtra)
# to get more information about surival regression
library(SurvRegCensCov)
# to predict usable covariates
library(rms)
```

## a)
We want to look at the lifetime distributions in the patients in the different treatment groups, \textit{standard} and \textit{test}.
```{r}
lungcancer.df <- as.data.frame(
  read.table("../data/TMA4275VeteranLungCancer.txt", 
             header = TRUE))
KM0 <- survfit(Surv(Y, C) ~ Treat,  
               type="kaplan-meier", 
               conf.type="log", 
               data=lungcancer.df)
ggsurvplot(KM0, data = lungcancer.df)
```
In the figure above the red curve is the survival distribution of the \textit{standard} treatment and the blue curve is the survival distribution of the \textit{test} treatment. From the plot we can see that the \textit{test} treatment is worse(steeper curve) in small lifetimes, but doesn't hit survival probability of $0$ for almost double the lifetime than for the \textit{standard} treatment. This tells us that \textit{test} is worse for '\textit{bad}' patients, but is better for '\textit{good}'. What a '\textit{bad}' or '\textit{good}' patient is, it's hard to say anything about yet, but might be possible after furter survival analysis and the inclusion of more covariates. The median and expected lifetimes of patients in the two groups is calculated in the code bellow and the results of which is shown in \ref{tab:treat}. From this we can see that median of the survival probaility is lower for the \textit{test} treatment, but the expected lifetime is alittle higher than for the \textit{standard} treatment. 
```{r}
treat.median <- summary(KM0)$table[, "median"]
treat.expected <- summary(KM0)$table[, "*rmean"]
table.treat <- data.frame(treatment = names(treat.median),
                           median = unname(treat.median),
                           expected = unname(treat.expected))
kable(table.treat,
      caption = "\\label{tab:treat}Estimated median and expected 
      lifetime by treatment") %>% 
  kable_styling(bootstrap_options = c("striped", "hover", 
                                      "condensed","responsive"),
                full_width = F,position = "center")
```
Now we do a similar assesment of a grouping by celltypes, which consist of the four different celltypes \textit{Squamous}($1$), \textit{small-cell}($2$), \textit{adeno}($3$) and \textit{large}($4$).   
```{r}
KM1 <- survfit(Surv(Y, C) ~ Cell,  
               type="kaplan-meier", 
               conf.type="log", 
               data=lungcancer.df)
ggsurvplot(KM1, data = lungcancer.df)
```
The figure above shows the survival distributions in the grouping by celltype. Overall $Cell = 2$ and $Cell = 3$ is the worst, and $Cell = 1$ looks like the best $Cell$ for lifetime. The code bellow calculates the median and expected lifetimes in the grouping by celltype, and the results are shown in the table \ref{tab:cell}.
```{r}
cell.median <- summary(KM1)$table[, "median"]
cell.expected <- summary(KM1)$table[, "*rmean"]
table.cell <- data.frame(treatment = names(cell.median),
                           median = unname(cell.median),
                           expected = unname(cell.expected))
kable(table.cell,
      caption = "\\label{tab:cell}Estimated median 
      and expected lifetime by cell type") %>% 
  kable_styling(bootstrap_options = c("striped", "hover",
                                      "condensed","responsive"),
                full_width = F,position = "center")
```
From this result we see that the $Cell=1$ has the heighest expected lifetime, followed by $Cell = 4$. But their roles are reversed for the median. For $Cell = 2$ and $Cell = 3$ they are almost equal in both accounts, but much lower lifetimes than the two others. The same conclusions as the one we drew from the figures. 
The reason the values of the median and expected lifetime not being consistently heigher than other group factors is because there are alot of data beeing either censored or dead the lower lifetimes. But the survival distribution might have data points at high lifetimes, and thereby have a long right tail, that will affect the expected lifetime alot. 


## b)
In this exercise we will perform a fit to the Weibull regression model, with $Treat$, $PS$, $Month$, $Age$ and $Prior$ as ordinary covariates, and $Cell$ as a factor represented by three dummy variables, $x_2$, $x_3$ and $x_4$. The result of which is shown in the code bellow, using $\texttt{survreg}$ from the R-package $\textbf{survival}$.
```{r}
wei.lung<-survreg(Surv(Y,C) ~ Treat + PS + Month+  Age + Prior + factor(Cell), data = lungcancer.df, dist = "weibull")
wei.lung
```
In the table \ref{tab:b} the defined indicators or dummy variables are shown. And we adapt this to our regression model.
```{r}
table.b <- data.frame(Levels.of.Cell = c(1,2,3,4), x2 =c(0,1,0,0), x3 = c(0,0,1,0), x4 = c(0,0,0,1))
kable(table.b,align=rep('c', 4),
      caption = "\\label{tab:b}Estimated median 
      and expected lifetime by cell type") %>% 
  kable_styling(bootstrap_options = c("striped", "hover",
                                      "condensed","responsive"),
                full_width = F,position = "center")
```
Given $x = \{x_1(\textrm{Treat}),x_2(\textrm{Cell=2}),x_3(\textrm{Cell=3}),x_5(\textrm{Cell=4}),x_5(\textrm{PS}),x_6(\textrm{Month}),x_7(\textrm{Age}),x_8(\textrm{Prior})\}$, the values of the covariates given some person, were the "\textit{dummy variables}" $x_2,x_3,x_4$ is set by table \ref{tab:b}, the model for log-lifetime $T$ is given by the expression, with corresponding covariates $\mathbf{x}$,
$$
  ln(T) = \beta_0 + \sigma\epsilon + \sum\limits_{i = 1}^8 \beta_ix_i.
$$
Where $\epsilon \sim G(0,\sigma)$, a gumbel distribution with shape $\sigma$. This gives us the hazard rate function
$$
  \begin{array}{rcl}
      z(t) & = & \alpha \cdot exp\{\beta_0 + \sum_{i=1}^8 \beta_ix_i\}^{-\alpha}\cdot t^{\alpha-1} \\
           & = & \alpha e^{-\alpha\beta_0}t^{\alpha-1}e^{-\alpha\cdot\sum_{i=1}^8 \beta_ix_i} \\
           & = & z_0(t) \cdot e^{-\alpha\cdot\sum_{i=1}^8 \beta_ix_i}
  \end{array}.
$$
Now we will calculate the estimated median lifetimes of patients with respectivly the same covariates as patient number $1$ and $19$, by this we look at the factor of $Cell$. Because it is the only covariate that would have discluded variables in the model based on its value. This means that for patient $1$ which has $\mathrm{Cell} = 1$ and patient $19$ which has $\mathrm{Cell} = 2$. In the function $\texttt{lifeWeil}$ the median lifetimes of a set of data is calculated using the estimated weibull model. 
```{r}
coef <- wei.lung$coefficients
lifeWeil <- function(data, coef){
  res = vector()
  for (i in seq(1,length(data[,1]))){
    x = c(1,data$Treat[i], data$PS[i], data$Month[i], data$Age[i], data$Prior[i],as.numeric(data$Cell[i]==2),
          as.numeric(data$Cell[i]==3),as.numeric(data$Cell[i]==4))
    res = c(res,exp(as.numeric(coef)%*%x))
  }
  return(median(sort(res)))
}
patient1 <- lungcancer.df[lungcancer.df$Cell == lungcancer.df$Cell[1],]
patient2 <- lungcancer.df[lungcancer.df$Cell == lungcancer.df$Cell[19],]
life1 <- lifeWeil(patient1,coef)
life2 <- lifeWeil(patient2,coef)
cat("Estimated median lifetime equal patient 1:", life1)
cat("Estimated median lifetime equal patient 19:", life2)
```

## c)
To look at the $p\textrm{-values}$ to determine significant effect of the covariates in our model, we can look at the output from the function $\texttt{summary.survreg}$.
```{r}
summary(wei.lung)
```
From the $p\textrm{-values}$ we can see that $PS$ and $Cell$ has significant effect. We can also look at the figure bellow to to get a more visual representation of the $p\textrm{-values}$.
```{r}
temp<-data.frame(covariates = names(summary(wei.lung)$table[,4]) ,pval = summary(wei.lung)$table[,4])
ggplot(temp) + 
  geom_point(aes(x = pval, y = covariates)) + 
  geom_vline(xintercept = 0.05, color = "firebrick") +
  scale_x_reverse()
```
The red line is the significance level of $0.05$. We observe that only $PS$ and two factors of $Cell$ is within acceptable $p\textrm{-values}$.

To look at the significant difference between the two treatments we use the function \texttt{ConvertWeibull} from the library \textbf{SurvRegCensCov}. From this function we get the HR(Hazard rate) or Risk of Death. We derive this by hand later in this exercise. 
```{r}
ConvertWeibull(wei.lung)$HR
```
Looking at the output and HR value for Treat, we can conclude that doing the \textit{test} treatment will increase the risk of death by approximately $28\%$.
The scale parameter of our Weibull distribution is $\alpha=$``r wei.lung$scale``, which is pretty close to an exponential disitribution of $\alpha = 1$. This means that an exponential disitribution might be a better model than our Weibull model. To determine this we need a better way to compare the models. One way of doing this is to calculate the AIC(Akaike's Information Criterion)-values of the two, and then choose the model with the smallest value. We will do this later in this exercise, but we recognize that the scale parameter is close to that of a exponential distribution.
Now we will take out the non-significant covariates in our model. This can be done by the $p\textrm{-values}$ found previously. Instead of just choosing we can use the function \texttt{fastbw} from the library \textbf{rms} to determine which covariates to use in our model based on $p\textrm{-values}$. If we choose a significance level of $5\%$ and then force our model to include Treatment we get the results shown in the code below, under '\textit{Factors in Final model}'. 
```{r}
psm.lung <- psm(Surv(Y,C)~Treat + PS + Month+  Age + Prior + factor(Cell), data = lungcancer.df,dist="weibull")
fastbw(psm.lung,rule = "p",sls = 0.05, force=c(1),type = "individual")
```
This means that the reduced model includes the covariates $Treat$, $PS$ and the factors of $Cell$. Which is also what we concluded earlier.
```{r}
wei.lung.new <- survreg(Surv(Y,C)~Treat + PS + factor(Cell), data = lungcancer.df, dist = "weibull")
summary(wei.lung.new)
```
Above we have done a Weibull fitting of the new reduced model. Looking at the scale $\alpha=$``r wei.lung.new$scale``, it is closer to that of a exponential disitribution. If we look at the $p\textrm{-value}$ plots we can see that only $Treat$ and $Cell = 4$ fall outside of the significance level of $0.05$.

```{r}
temp2<-data.frame(covariates = names(summary(wei.lung.new)$table[,4]) ,pval = summary(wei.lung.new)$table[,4])
ggplot(temp2) + 
  geom_point(aes(x = pval, y = covariates)) + 
  geom_vline(xintercept = 0.05, color = "firebrick") +
  scale_x_reverse()
```
If we again look at the risk of death(HR) values from this model fit, the significant difference between the two treatments have dropped alittle, but the risk of death for the \textit{test} treatment is $25\%$ higher than for the \textit{standard} treatment.
```{r}
ConvertWeibull(wei.lung.new,conf.level = 0.95)$HR
```


## d)
The Cox-model for the reduced model we found in c) is given by the equation
$$ 
  z(t;\mathbf{x}) = z_0(t)e^{\beta_1x_1 + \beta_2x_1 + \beta_3x_3+ \beta_4x_4 + \beta_5x_5},
$$
where $z_0(t)$ is any non-parametric hazard rate function and is unkown in the cox model. The reduced Weibull model for hazard rate is given by the equation
$$
  z(t;\mathbf{x}) = \alpha e^{-\alpha\beta_0}\cdot t^{\alpha-1}\cdot e^{-\alpha(\beta_1x_1 + \beta_2x_2 + \beta_3x_3 + \beta_4x_4 + \beta_5x_5)},
$$
and the $z_0(t)$ function is known. Now we perform a cox regression on our data. 
```{r}
fit.coxph <- coxph(Surv(Y,C) ~ Treat + PS+ factor(Cell), data = lungcancer.df)
fit.coxph
```
If we look at the relative risk of this Cox-model. The relative risk is found by increasing the value of one covariate by $1$ and keeping all the other covariates constant. Then calculating the quotient of the new and previous hazard rates, $z(t,x_i+1)$, $z(t,x_i)$, one gets the relative risk $RR$. Given by the equation
$$
 RR = e^{\beta_i},
$$
with $i$ being the covariate of interest, where in our case $i \in \{1,2...,5\}$. This means that from the output of our Cox-regression above we look at the values of $exp(coef)$, which is the relative risk factor. The highest relative risk is from $\mathrm{Cell} = 3$. The risk of death is increased with $26\%$ if one uses the \textit{test} treatment istead of the \textit{standard} treatment. If one increases the \textit{Performance status} by $1$ the change is not that significant. But since $PS$ has a large range, $x_5\in[10,90]$, the change is more significant than what the relative risk factor is suggesting. 

## e)
In this exercise we will look at the interaction between two covariates, $Treat$ and $Cell$, while keeping $Cell$ as a factor. We are still using table \ref{tab:b} for the values of $x_2$,$x_3$ and $x_4$. With the interaction of the covariates, we introduce three new variables $z_2 = x_1\cdot x_2$, $z_3 = x_1\cdot x_3$ and $z_4 = x_1\cdot x_3$. We will use the same covariates as in the reduced model from c), $Treat$, $PS$, and $Cell$, and we use weibull regression to determine the $\beta_i$- values.
```{r}
wei.lung.int<- survreg(Surv(Y,C) ~ Treat + PS+ factor(Cell) + Treat*factor(Cell), data = lungcancer.df, dist = "weibull")
summary(wei.lung.int)
```
As hinted in the exercise paper the only interaction which has $p\mathrm{-value}$ lower than a significance level of 0.05 is $z_2$.

We will now to calculate the estimated relative risk of patients with $Cell = 2$ in the \textit{standard} treatment, compared to the patient with the \textit{test} treatment. First we find a general formula for the Relative Risk($RR$) given the $Cell$. Let $\mathbf{x} = \{x_1,x_2,x_3,x_4,x_5,x_1\cdot x_2,x_1\cdot x_3,x_1\cdot x_4\}$ and $\mathbf{x} = \{x_1+1,x_2,x_3,x_4,x_5,(x_1+1)\cdot x_2,(x_1+1)\cdot x_3,(x_1+1)\cdot x_4\}$, then the relative risk is given by the equation
$$
\begin{array}{rcl}
  RR_{\textrm{weib}}(x_1|Cell=j;j\in\{2,3,4\}) & = & \frac{z(t;\mathbf{x}_2|Cell=j)}{z(t;\mathbf{x}_1|Cell=j)}\vspace{1em}\\
                               & = & \frac{z_0(t)\cdot \exp\{-\alpha\cdot(\beta_1(x_1+1) + \beta_jx_j + \beta_5x_5 + \beta_{j+4}(x_1+1)x_j)\}}{z_0(t)\cdot \exp\{-\alpha\cdot(\beta_1x_1+ \beta_jx_j + \beta_5x_5 + \beta_{j+4}x_1x_j)\}}\vspace{1em}\\ 
                               & = & \exp\{-\alpha\cdot(\beta_1 + \beta_{j+4})\}
\end{array}
$$
If the $Cell = 1$, the expression simply becomes
$$
RR(x_1|Cell=1) = \exp\{-\alpha\cdot\beta_1\}
$$
From this knowledge we can create a function in \textbf{R} that calculates the relative risk of all the different $Cell$ types respectivly. This is done in the function \texttt{relativeRisk} and the result of which is shown in table \ref{tab:e}.
```{r}
relativeRisk <- function(coef,Cell,scale){
  if (Cell > 1){
    return(exp(-scale*(coef[2] + coef[Cell+5])))
  }
  return(exp(-scale*coef[2]))
}
coef.int <- wei.lung.int$coefficients
scale.int <- wei.lung.int$scale
table.e <- data.frame(Cell = c(1,2,3,4), 
                      RR = c(relativeRisk(coef.int,Cell = 1,scale.int),
                             relativeRisk(coef.int,Cell = 2,scale.int),
                             relativeRisk(coef.int,Cell = 3,scale.int),
                             relativeRisk(coef.int,Cell = 4,scale.int)))
kable(table.e,align=rep('c', 4),
      caption = "\\label{tab:e}Estimated Relative Risk of 
      \\textit{test} treatment\n compared to \\textit{standard} 
      treatment in the different \\textit{Cell}-types.") %>% 
  kable_styling(bootstrap_options = c("striped", "hover",
                                      "condensed","responsive"),
                full_width = F,position = "center")
```
From the results in table \ref{tab:e} we can see that for patients with $Cell$-type $1$ it is approximately $25\%$ lower risk of death using the \textit{test} treatment. For patients with $Cell$-type $2$ it is approximately $87\%$ higher risk of death using the \textit{test} treatment. Further on for patients with $Cell$-type $3$, it is approximately $19\%$ lower risk of death using the \textit{test} treatment. And last for patients with $Cell$-type $4$ it is approximately $48\%$ higher risk of death using the \textit{test} treatment. Thereby we can conclude that patients with $Cell$-type $1$ should take \textit{test} treatment, patients with $Cell$-type $2$ should take the \textit{standard} treatment, patients with $Cell$-type $3$ should take \textit{test} treatment and patients of $Cell$-type $4$ should take the \textit{standard} treatment. 


## f)
From the model fit in the final model in exercise e), we used a Weibull regression with reduced amount of covariates from what we started out with. We used $Treat$, $Cell$ as factor, $PS$ , and  $Treat \times Cell$ with $Cell$ as factor. We will now try to fit different distributions to the data using the same covariates as in e).
We tested the exponential distribution yielding:
```{r}
exp.lung<- survreg(Surv(Y,C)~Treat + PS + factor(Cell)+ factor(Cell)*Treat, data = lungcancer.df, dist = "exponential")
summary(exp.lung)
```
The gaussian distribution:
```{r}
gauss.lung<- survreg(Surv(Y,C)~Treat + PS + factor(Cell)+ factor(Cell)*Treat, data = lungcancer.df, dist = "gaussian")
summary(gauss.lung)
```
The logistic distribution:
```{r}
log.lung<- survreg(Surv(Y,C)~Treat + PS + factor(Cell)+ factor(Cell)*Treat, data = lungcancer.df, dist = "logistic")
summary(log.lung)
```
The lognormal distribution:
```{r}
lognorm.lung<- survreg(Surv(Y,C)~Treat + PS + factor(Cell)+ factor(Cell)*Treat, data = lungcancer.df, dist = "lognormal")
summary(lognorm.lung)
```
The log-logistic distribution:
```{r}
loglog.lung<- survreg(Surv(Y,C)~Treat + PS + factor(Cell)+ factor(Cell)*Treat, data = lungcancer.df, dist = "loglogistic")
summary(loglog.lung)
```
From the loglikelihood and the number of parameters $k$ in the different distributions we can caluclate the $AIC$(Akaike's Information Criterion) by the equation
$$
  AIC = -2\cdot log L + 2\cdot k
$$
```{r}
AIC.wei <- -2*wei.lung.int$loglik[1]+2*summary(wei.lung.int)$df
AIC.exp <- -2*exp.lung$loglik[1]+2*(1 - pchisq(summary(exp.lung)$chi, (summary(exp.lung)$df-summary(exp.lung)$idf)))
AIC.gauss <- -2*gauss.lung$loglik[1]+2*(1 - pchisq(summary(gauss.lung)$chi, (summary(gauss.lung)$df-summary(gauss.lung)$idf)))
AIC.log <- -2*log.lung$loglik[1]+2*(1 - pchisq(summary(log.lung)$chi, (summary(log.lung)$df-summary(log.lung)$idf)))
AIC.lognorm <- -2*lognorm.lung$loglik[1]+2*(1 - pchisq(summary(lognorm.lung)$chi, (summary(lognorm.lung)$df-summary(lognorm.lung)$idf)))
AIC.loglog <- -2*loglog.lung$loglik[1]+2*(1 - pchisq(summary(loglog.lung)$chi, (summary(loglog.lung)$df-summary(loglog.lung)$idf)))

table.f <- data.frame(Distribution = c("weibull","exponential","gaussian","logistic","lognormal","loglogistic"), 
                      AIC = c(AIC.wei,AIC.exp,AIC.gauss,AIC.log,AIC.lognorm,AIC.loglog))

kable(table.f,align=rep('c', 4),
      caption = "\\label{tab:f}Estimated Relative Risk of 
      \\textit{test} treatment\n compared to \\textit{standard} 
      treatment in the different \\textit{Cell}-types.") %>% 
  kable_styling(bootstrap_options = c("striped", "hover",
                                      "condensed","responsive"),
                full_width = F,position = "center")
```